

The first critical phase of this project was to replicate the original RISCV-Matrix execution environment and rebuild the benchmark suite under the current toolchain. The old speed and accuracy results of RISCV-Matrix were generated using an older cross-compiler toolchain and test suite configuration. This section compares those original results, designated the Grace baseline (commit \texttt{113f4c66}, December 2022), against the current, optimized pipeline (April 2026). All cycle counts were measured on the Spike RISC-V ISA simulator with a vector length of VLEN=1024.

Next, the focus was to benchmark the ts-traversal translation pipeline with various MATLAB and Octave tests for correctness and speed. Finally, some non-vectorized functions from RISCV-Matrix were vectorized and mapped onto the ts-traversal pipeline.

\section{Additional Vectorization}

The Grace baseline shipped vectorized kernels for the full binary, arithmetic, and comparison set. The work reported here extends the vectorized set by adding \texttt{\_vec} entry points for \texttt{transposeM}, \texttt{ctransposeM}, \texttt{meanM}, \texttt{varM}, \texttt{minM}, \texttt{maxM}, \texttt{minV}, \texttt{maxV}, \texttt{stdM}, and \texttt{popstdM}. These additions now appear as new rows in Tables~\ref{tab:ratio-unary} and~\ref{tab:cycles-unary}.

The ten new entry points land in April 2026. Accuracy tests pass for all of them at every size from $N=1$ to $N=1{,}050{,}625$ except for two tests at 127x127 matrices: \texttt{varM}, \texttt{stdM}, and \texttt{mtimesM} (a vectorized kernel written before this research) have their vec and ref outputs off by a factor of at most $8.941 \times 10^{-8}$, $2.444 \times 10^{-9}$ and $1.714 \times 10^{-7}$ respectively.

The ten new functions have not all been vectorized for all possible dimensions and types. For \texttt{transposeM} and \texttt{ctransposeM}, all 2D types have been vectorized while all 1D inputs fall back to the reference implementation. For \texttt{meanM} and \texttt{varM}, 1D doubles and all 2D types have been vectorized. 
For \texttt{minM/V} and \texttt{maxM/V}, all but 1D and 2D complex inputs have been vectorized due to the need for \texttt{carg} for complex comparison. \texttt{stdM} and \texttt{popstdM} inherit the exact same constraints as \texttt{varM}, remaining fully vectorized only for 1D doubles and all 2D matrix types.
% For \texttt{stdM} and \texttt{popstdM}, the variance is computed through the existing \texttt{varM\_vec}/\texttt{popvarM} dispatch, then an in-place \texttt{vfsqrt\_v\_f64m1} loop replaces the reference path's \texttt{scalarpowerM(\_, 0.5, DOUBLE)} (which trampolines through libc \texttt{pow}); INTEGER and DOUBLE only, COMPLEX falls back to ref because principal-branch complex sqrt is non-trivial in RVV.

\section{Total Runtime Comparison}

Table~\ref{tab:total-runtime} details the total runtime comparison, displaying the specific cycle counts and the corresponding percentage changes in execution speed.

\input{tables/table_total_runtime}

An analysis of Table~\ref{tab:total-runtime} reveals several insights about the current state of the compilation pipeline. Across the entire test suite, the overall execution time decreased by approximately 3.9 percent, dropping from 419.50 Giga-cycles to 403.01 Giga-cycles. RISC-V GNU compiler collection (GCC) updates played a crucial part in these improvements. The distribution of these performance gains is non-uniform. For instance, the majority of the standard arithmetic and logical operations experienced moderate speedups ranging from 10-20 percent. All four core arithmetic tests dropped by 10-17\%, consistent with the GCC toolchain improvements. This demonstrates that the GCC updates benefit both vectorized and scalar code paths across these operations.

The four comparison tests utilizing Octave-mode semantics--\texttt{octave\_ge}, \texttt{octave\_gt}, \texttt{octave\_le}, and \texttt{octave\_lt}--showed negligible improvement ($< 1\%$). These tests dominate the total runtime across the test suite at approximately 51 Giga-cycles each (versus $\sim$3.6 Giga-cycles for their MATLAB-mode counterparts) because Octave comparison semantics for complex numbers require computing the magnitude ($|z| = \sqrt{\text{re}^2 + \text{im}^2}$) for both operands and, when magnitudes are equal, resolving ties via the phase angle ($\text{atan2}(\text{im}, \text{re})$) \cite{octave_comparison_ops}. The phase-angle path (\texttt{arg}) required by Octave currently executes as a scalar libc arg call because it was never mapped to RVV intrinsics, making it the dominant cost. In contrast, MATLAB-mode comparisons only compare the real parts of complex numbers, requiring a single vector comparison instruction.

\subsection{The \texttt{divide} Anomaly}

The \texttt{divide} test is the only operation that exhibited a total runtime increase ($+3.0\%$, from 5.47 to 5.64 Gcycles). This regression is localized to complex division reference counts -- the reference cycle counts for complex-involving type combinations approximately doubled between Grace and current (e.g., \texttt{complex$\times$complex ref}: 36.1M $\rightarrow$ 61.3M cycles), while the vectorized cycle counts remained stable ($\sim$8M cycles). This suggests that the updated GCC generates less efficient scalar code for complex division.

In contrast, the \texttt{int$\times$int} reference for \texttt{divide} dropped from 14.4M to 4.8M ($-67\%$), indicating that the compiler dramatically improved integer-only scalar division. This shows that the same toolchain update can simultaneously upgrade and degrade performance across different data type combinations, even within a single function.

\section{Vectorization Effectiveness}

Further analysis was conducted to evaluate the vectorization implementation relative to scalar execution across different variable types. Tables~\ref{tab:ratio-unary} and~\ref{tab:ratio-binary-diag} present the vector-to-reference execution ratios at the largest tested matrix size ($1025 \times 1025$). A ratio less than 1.0 indicates that the vectorized implementation is faster than the scalar reference. The tables compare Grace and current ratios side-by-side.

Note on type combinations: for binary operations (e.g., \texttt{plusM}, \texttt{divideM}), the test suite evaluates every combination of input types. The notation ``int $\times$ double'' means the left operand matrix contains integers and the right operand matrix contains doubles. The result type is determined by C type promotion rules.

\input{tables/table_ratio_unary}

\input{tables/table_ratio_binary_diag}

A comprehensive breakdown of all nine type combinations for each binary operation is provided in Table~\ref{tab:ratio-binary-full} in the Appendix.

\subsection{Ratio Shifts Due to Compiler Improvements}

While the vectorized implementations maintained approximately constant cycle counts between Grace and current (for the previously existing vectorized workloads), several vec/ref ratios increased significantly--meaning the scalar \emph{reference} code got faster, narrowing the vectorization advantage. Three distinct patterns were identified:

\paragraph{Pattern 1: Integer memory operations.} The integer reference cycle counts dropped uniformly for the \texttt{ceil}, \texttt{floor}, \texttt{round}, \texttt{full\_duplicate}, and \texttt{ones} tests. Many of these tests saw their ratios jump by $\sim 29\%$ across int and double types. This is likely because GCC optimized memory load/store type operations since the Grace era.

\paragraph{Pattern 2: Logical operations on non-integer types.} A similar ref speedup was observed for logical operations on double and complex types. The cycle count for \texttt{not} and \texttt{true} decreased by $\sim$65\% (e.g., \texttt{not} double ref: 10.3M $\rightarrow$ 3.6M) while the vectorized counts stayed at $\sim$3.1M. This caused the double/complex ratios to jump from $\sim$0.30 to $\sim$0.85. The GCC compiler now seems to optimize these operations more aggressively.

\paragraph{Pattern 3: Binary operations.} Several binary operations exhibited compiler-driven ref improvements. \texttt{matlab\_ge} through \texttt{matlab\_lt} (dbl$\times$dbl), and \texttt{octave\_ge} through \texttt{octave\_lt} (dbl$\times$dbl) all showed reference speedups of 3--5$\times$, pushing their ratios from $\sim$0.25 to $\sim$0.83. The full details on binary operation cycles and ratios are available in the Appendix (Table~\ref{tab:ratio-binary-full}).

In all cases, the vectorized code did not get slower--the reference code simply got faster. The absolute performance of the library improved or stayed steady, but the relative advantage of explicit vectorization narrowed for these specific combinations.

\subsection{The \texttt{isEqual} Anomaly}

The \texttt{isEqual} test consistently returns a vec/ref ratio greater than 1.0 for complex type combinations (e.g., complex$\times$complex: 1.172 in Grace, 1.121 in current), indicating that the vectorized implementation is slower than scalar. This turns out to be caused by the test harness, not the implementation. The test harness short-circuits the comparison, so the vectorized code never gets to the expensive part of the comparison.

The evidence shows that the cycle counts for \texttt{isEqual} complex$\times$complex remain \textbf{exactly 75 cycles (vec) and 64 cycles (ref) at every single matrix size tested}--from $N=1$ through $N=1{,}050{,}625$:

\begin{table}[H]
\centering
\caption{The \texttt{isEqual} complex$\times$complex cycle counts remain constant across all matrix sizes, proving short-circuit evaluation.}
\label{tab:isequal-evidence}
\small
\begin{tabular}{r r r r}
\toprule
Size $N$ & Vec (cycles) & Ref (cycles) & Ratio \\
\midrule
1 & 75 & 64 & 1.172 \\
9 & 75 & 64 & 1.172 \\
16,384 & 75 & 64 & 1.172 \\
1,050,625 & 75 & 64 & 1.172 \\
\bottomrule
\end{tabular}
\end{table}

If the implementations were actually traversing the data, the cycle counts would scale linearly with $N$. The lack of linear scaling shows that both vec and ref implementations exit after comparing only the \emph{first element}. An examination of the test data generation confirmed that the \texttt{isEqual} test compares matrices populated with entirely divergent data streams, causing an immediate short-circuit mismatch.

The 17\% performance gap (75 vs.\ 64 cycles) is attributable to \emph{vector setup overhead}. \texttt{isEqual\_vec} executes vector length configuration and stride-loading instructions before performing comparisons. However, the scalar path simply loads and compares two values, so it does not have the same overhead cost. In workloads where equality must be verified across a full traversal, the vectorized implementation vastly outperforms scalar execution, as can be seen in the complex$\times$double combination achieving a ratio of 0.015 (66$\times$ speedup).

\section{Toolchain-Induced Vectorization Regressions}

While the preceding section described cases where the vec/ref ratio \emph{increased but remained below 1.0} (vec still faster), three specific type combinations crossed the 1.0 threshold entirely--meaning the scalar reference now outperforms the vectorized implementation:

\begin{table}[H]
\centering
\caption{Type combinations where the scalar reference became faster than the vectorized implementation in the current toolchain.}
\label{tab:regressions}
\small
\begin{tabular}{l l r r r r}
\toprule
Test & Type Combo & \multicolumn{2}{c}{Grace} & \multicolumn{2}{c}{Current} \\
 & & Ratio & Ref (cycles) & Ratio & Ref (cycles) \\
\midrule
and & dbl $\times$ dbl & 0.493 & 10.9M & 2.678 & 2.0M \\
and & cpx $\times$ dbl & 0.408 & 13.3M & 2.464 & 2.2M \\
or & cpx $\times$ dbl & 0.477 & 11.4M & 2.392 & 2.2M \\
\bottomrule
\end{tabular}
\end{table}

In all three cases, the vectorized cycle counts remained approximately constant ($\sim$5.3M for \texttt{and}, $\sim$5.4M for \texttt{or}). The regressions are entirely driven by GCC's updated optimizer reducing the reference cycle counts by a factor of five to six. It is expected that the ``scalar'' reference is no longer scalar: GCC 15's RISC-V backend auto-vectorizes it, and the resulting code beats the library's hand-written kernel.
% because it fuses the compare and mask steps that the intrinsic version keeps separate.

This has two implications. First, compiler evolution can erode the benefits of manual vectorization. Second, the hand-written kernels here use \texttt{LMUL=8}, which the RISCV-Matrix README warns will ``cause extra loads and stores to be called implicitly'' when the kernel uses more than a handful of live vector registers \cite{riscvmatrix}. Investigating the LMUL choice is therefore a suggested ratio recovery path for future work.

\section{Absolute Cycle Counts}

Table~\ref{tab:cycles-unary} presents the absolute vectorized and reference cycle counts at the largest matrix size for both Grace and current environments to display the raw performance data.

\input{tables/table_cycles_unary}

These raw counts confirm that vectorized implementations maintained or slightly improved their absolute performance, while compiler improvements to the scalar reference code drove all observed ratio shifts.

\section{TS-traversal Compiler}

Beyond the resolution of benchmarking anomalies, significant debugging efforts have focused on addressing algorithmic failures within the ts-traversal compiler phase. When the project was inherited, the TypeScript traversal would emit C that compiled but failed on several fundamental constructs: higher-dimensional indexing (3-D and 4-D array subscripts), various standalone issues that propagated across multiple functions like off-by-one for loop iteration, missing column vector support, missing function mapping, and row major vs. column major representation across MATLAB and C. There were also some tests that were producing different results across the MATLAB and generated C files due to a difference in the type of BLAS (Basic Linear Algebra Subprogram) libraries used between MATLAB (OPENBLAS, requires POSIX) and our bare metal RISC-V binaries (LAPACK, RISCVBLAS). The tolerance level has now been set to $5 \times 10^{-10}$ for both absolute and mantissa level checking. After various fixes, each of these paths now take a representative MATLAB script through translation, compilation with \texttt{riscv64-unknown-elf-gcc} 15.2.0, and execution on Spike without harness-level failure.

To measure the translated output, a \texttt{SPEED\_*} directory family was added to the generated-code corpus that exercises the translated kernels at three problem sizes (\texttt{small} $2 \times 2$, \texttt{medium} $100 \times 100$, \texttt{big} $1000 \times 1000$) across nine operations: \texttt{addition}, \texttt{eigen}, \texttt{fourier}, \texttt{inverse}, \texttt{multiplication}, \texttt{power\_complex}, \texttt{power\_int}, \texttt{power\_double}, and \texttt{transpose}. There also exist 61 other smaller test kernels such as \texttt{reindex}, \texttt{fourier}, and \texttt{filterM}.  Each directory is built by a series of \texttt{make} commands, which regenerates the C source from the \texttt{.m} input, cross-compiles the binary, and records the Spike cycle count.

To measure the effect of this semester's translation-layer and library changes, the \texttt{make timediff} target replays every test through Spike and compares the retired-instruction count against a baseline captured at the 68-tests-passing checkpoint (before the current round of bug fixes began). The list of fixes discusses previously brought up the passing test count to 84/88.  The 4 remaining comparison failures (\texttt{class\_test}, \texttt{octaveIncludes}, \texttt{octaveTimer}, and \texttt{octave\_syntax}) are due to pre-existing pipeline constraints. The Octave tests fail because they contain Octave-specific syntax configurations (such as \texttt{\%source} directives and \texttt{endfunction}) which are not formally supported by the \texttt{tree-sitter-matlab} grammar, causing the pipeline to generate invalid C code. Similarly, \texttt{class\_test} fails to generate evaluable C code because the transpiler currently lacks mappings for the \texttt{EEGLAB mmo} object structure. For \texttt{octaveTimer}, while the generated code successfully compiles and executes, the comparison fails due to system timer disparities; the C Standard Library's \texttt{time()} function provides 1-second resolution whereas MATLAB's \texttt{tic}/\texttt{toc} offers 1-microsecond precision, leading to a fundamental difference in timing.

In terms of cycle count improvements, Table~\ref{tab:ts-timediff} summarizes the results.

\input{tables/table_ts_timediff}

The observed speed improvement was less than initially expected due to the fact that ts-traversal changes and tests focused on improving and checking correctness and not performance. There was a hypothesis that the further vectorization of some functions like \texttt{transposeM} and \texttt{varM} would lead to improvements for their respective test cases. While most SPEED tests did not show improvement, \texttt{transposeM} did show notable improvement -- $0.93\%$ on \texttt{SPEED\_big\_transpose} and $26.75\%$ on \texttt{SPEED\_medium\_transpose}. While the numbers seem tame, they tell a bigger story of improvement due to the \texttt{transposeM} calls in these test cases being made 1 and 100 times respectively. This suggests that with further vectorization, even if a heavy math function is called only a few times, having it vectorized enables quicker transition into desired idle states for patient safety. The main reason for the tests that regressed in their cycle counts is the print statement heavy nature of the tests. The definition of \texttt{print} was changed to print more precise values in order to have accurate tolerance checking and get more tests passing, causing significant slowdown for these tests. The lack of change in the \texttt{SPEED\_big\_fourier} test case supports the \texttt{print} statement hypothesis, as it has no \texttt{print} statements and thus was unaffected by the change.

\subsection{C-Octave Suite Residuals}

There is a suite of Octave and C tests that mirror the same kernels tested in the MATLAB-vs-C comparison, only using Octave instead of MATLAB, which causes certain behavioral differences to be discussed in this section. The same tolerance gate (\texttt{tolerant\_diff.py}, $\text{MANTISSA\_TOL} = \text{ABS\_TOL} = 5 \times 10^{-10}$, $\text{EXP\_JITTER} = 1$) is now wired into \texttt{tests\_C\_Octave/Makefile}'s \texttt{compare\_outputs} target so the same numerical contract governs both surfaces. Following recent fixes to display handling, the number of passing tests went from 66 to 71/79. The 8 still-failing tests fail due to \emph{FFT library divergence} or \emph{error handling behavior differences}. 
